\texttt{nuance}: Detection of planetary transits in the presence of correlated noise

### 2023-01-20 simple/latex/ms.tex
simulation appendix started

% Define document class
\documentclass{aastex631}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{bm}

@@

Throughout this paper, we simulate stellar variability (and later model it) thanks to Gaussian Processes (GP), employing the physically-motivated stochastically-driven damped simple harmonic oscillator kernel (SHO) presented in \citealt{celerite1} (see Annexe ...). We generate stellar variability thanks to the \texttt{tinygp}\footnote{\href{https://github.com/dfm/tinygp}{https://github.com/dfm/tinygp}} Python package, providing an implententation of the scalable GP method from \citealt{celerite2} powered by \texttt{JAX}\footnote{\href{https://github.com/google/jax}{https://github.com/google/jax}}.

@@

During the \textit{linear search}, the goal is to compute the likelihood $p(\bm{f} \vert T , D, \Delta)$ of our data given a single non-periodic transit signal of epoch $T$, duration $D$ and depth $\Delta$, for a grid of epochs, durations and depths.
\\\\
To account for correlated noises, we model the light curve $f$ as being drawn from a Gaussian Process such that
$$\bm{f} \sim \mathcal{N}(\bm{w X}, \bm{\Sigma})$$
with mean $\bm{wX}$ (i.e. a linear model of the M explanatory variables with coefficients $\bm{w}$) and covariance $\bm{\Sigma}$. To account for the presence of a single non-periodic transit of epoch $T$ and duration $D$, we compute and append its signal as the last column of the design matrix $\bm{X}$, using the simple transit model from \citealt{protopapas} with a unitary depth. This way, the transit signal is part of the linear model and its depth $\Delta$ can be solved linearly. Under this assumption, the log-likelihood of the data given a single non-periodic transit is analytical
\begin{equation} \label{eq:linear_search_ll}
    \ln p(\bm{f} \vert I) = -\frac{1}{2}(\bm{f}-\bm{wX})^T\bm{\Sigma}^{-1}(\bm{f}-\bm{wX}) -  \frac{1}{2}\vert\bm{\Sigma}\vert - \frac{N}{2}\ln 2
\end{equation}
and can be computed on a grid of epochs and durations, the transit depth being linearly solved for any $(T, D)$. This \textit{linear search} lead to the set of  log-likelihoods\footnote{This notation omits the vector $\bm{w}_{i,j}$ (except its last value $\Delta_{ij}$) as it is linearly solved for any given value of $(T_i, F_j)$ and irrelevant in what follows.}
$$\ln\mathcal{L} = \{\ln p(\bm{f} \vert T_i ,D_j, \Delta_{ij})\}_{i, j}$$
were $\Delta_{ij}$ is the depth linearly solved for a given $(D_i, T_j)$. \autoref{fig:linear_search} shows this likelihood grid computed for the simulated dataset shown in \autoref{fig:principle_dataset}, using the same Gaussian Process and design matrix $\bm{X}$ used to simulate the data (see annexe).

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.75\linewidth]{../figures/principle_linear_search.pdf}
        \caption{Principle and output of the \textit{linear search}. First, a set of durations and depths $\set{T_i, D_j}_{i,j}$ is generated. For each pair of index $(i,j)$, the log likelihood $\ln p(\bm{f} \vert T_i ,D_j, \Delta_{ij})$ is computed, using the parameters from \autoref{eq:ls} and the expression of \autoref{eq:linear_search_ll}. This process yields the grid of log-likelihoods $\ln\mathcal{L}$ (bottom plot), as well as the $\set{\Delta_{ij}, \sigma_{ij}}_{i, j}$ transit depths and errors.}
        \label{fig:linear_search}
    \end{centering}
\end{figure}

To prepare the next step, the corresponding depths $\Delta_{ij}$ linearly solved for any $(T_i ,D_j)$ are conserved, as well as their associated errors $\sigma_{ij}$ obtained through the least-square solution
\begin{equation}\label{eq:ls}
    \begin{gathered}
        \bm{w} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}\bm{X}^T\bm{\Sigma}^{-1}\bm{f} \\
        \bm{\sigma} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}
    \end{gathered}
\end{equation}    
where $\Delta_{ij} = \bm{w}_M$ and $\sigma_{ij} = \bm{\sigma}_{MM}$

@@

We then need to combine the likelihoods computed from the \textit{linear search} to obtain
\begin{equation}\label{eq:goal}
    p(\bm{f} \vert P, T_0 , D, \Delta)
\end{equation}
For a given transit duration $D$, any combination of $(P, T_0)$ leads to K truly independent transits, for which it is tempting to write
\begin{equation}\label{eq:attempt}
    p(\bm{f} \vert P, T_0 ,D, \Delta) = \prod_k^K p(\bm{f} \vert T_k, D, \Delta_k)
\end{equation}
where $\set{T_k}_k$ are the epochs matching $(T_0, P)$ and $\set{\Delta_k}_k$ the corresponding depths. So that
$$
    \ln p(\bm{f} \vert P, T_0 ,D, \Delta) = \sum_k^K \ln \mathcal{L}_k
$$
This is the joint log-likelihood of transits belonging to the same periodic signal but with varying depths  $\set{\Delta_k}_k$. However, individual transits from a periodic signal cannot be considered independent, and should instead share a common transit depth $\Delta$. We show in \autoref{combining_transits} that there is an analytical expression for the joint log-likelihood of K individual transits with depths and errors $\set{\Delta_k, \sigma_k}_k$ assuming a common depth $\Delta$:
\begin{equation}\label{eq:comb}
    \begin{gathered}
        \ln p(\bm{f} \vert P, T_0 ,D, \Delta) =  \sum_{k}^K \ln \mathcal{L}_k  - \frac{K}{2}\ln(2\pi) - \frac{1}{2}\sum_k^K \left( \frac{\left(\Delta_{k} -
        \bar\Delta\right)^{2}}{\sigma_k^{2} + \sigma^{2}} + \ln{\left(\sigma_k^{2} + \sigma^{2} \right)} \right) \\
        \text{with} \quad  \frac{1}{\sigma^2} = \sum_k^K \frac{1}{\sigma_k^2} \quad \text{and} \quad
        \Delta = \sigma^2 \sum_k^K {\frac{\Delta_k}{\sigma_k^2}}
    \end{gathered}
\end{equation}
While this gives a solution for \autoref{eq:goal}\footnote{Different from the one found in \cite{foreman2016}}, the individual epochs matching $T_0$ and $P$ are not necessarily available in the grid of epochs $\set{T_k}_k$. In \cite{foreman2016} this is solved by using the nearest neighbors in the epochs grid. Instead, to allow the efficient matrix computation of \autoref{eq:comb}, we interpolate the likelihood grid from $\set{T_i}_i$ to a common grid of transit phases $\set{\phi_i}_i$, leading to the \textit{periodic search} log-likelihood
$$\ln\mathcal{P}(P) = \set{\ln p(\bm{f} \vert P, \phi, D)}_{i,j}$$
shown for few periods in \autoref{fig:periodic_search}.b, a notation where $\Delta_{ij}$ is omitted since being interpolated from the \textit{linear search} using $\phi_i$ and $D_j$, and $T_0 = 0$.

\begin{figure}[H]
    \begin{centering}
        \gridline{\fig{../figures/principle_periodic_0.pdf}{0.7\linewidth}{\vspace{-0.5cm}(a) $\ln \mathcal{L}$}}
        \vspace{-0.6cm}
        \gridline{\fig{../figures/principle_periodic_1.pdf}{0.6\linewidth}{\vspace{-0.5cm}(b) $\mathcal{P}(P)$}}
        \caption{On the example dataset, this figure shows how the \textit{periodic search} works at different periods $P$ (including the true period $P = 0.7$ days). Given $P$ and $T_0=0$, the log-likelihood $\ln\mathcal{L}$ shown in (a) is phase-folded and interpolated onto a common grid of phases shown in (b). As an example, the white lines in (a) mark the edges of each fold for a period of $P=0.5$ days, and the white crosses the epochs $\set{T_k}_{k\in\mathbb{K}}$ matching a particular phase in the grid (reported in (a) and (b)) on which the corresponding $\set{\ln \mathcal{L}_k}_{k\in\mathbb{K}}$ are interpolated and combined using \autoref{eq:comb}. To allow the use of efficient matrix computations, this is done for every duration $\set{D_i}_i$ so that $\mathcal{P}$ is computed on the full grid $\set{D_i, \phi_i}_{i, j}$ at once. We understand from plots in (a) that a different choice of epoch $T_0$ may only shift the results in phase but do not affect the values of $\mathcal{P}$. For this reason, computing $\mathcal{P}$ for $T_0=0$ is sufficient. We also notice how the maximum value of $\mathcal{P}(P=0.35)$ is lower than for $P=0.7$ days, a result of combining the log-likelihoods using \autoref{eq:comb} instead of \autoref{eq:attempt}, in favor of individual transits matching a common depth $\Delta$.}
        \label{fig:periodic_search}
    \end{centering}
\end{figure}

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.62\linewidth]{../figures/principle_Q.pdf}
        \caption{}
        \label{fig:periodogram}
    \end{centering}
\end{figure}

@@

\newpage
\appendix
\section{Light curve simulations}\label{app_principle_simulations}
\section{Combining transits}\label{combining_transits}

>>>>

% Define document class
\documentclass{aastex631}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{bm}
\usepackage{xcolor}

@@

Throughout this paper, we simulate stellar variability (and later model it) thanks to Gaussian Processes (GP), employing the physically-motivated stochastically-driven damped simple harmonic oscillator kernel (SHO) presented in \citealt{celerite1}. Extensive details about the following simulations can be found in \autoref{app_principle_simulations}.

@@

During the \textit{linear search}, the goal is to compute the likelihood $p(\bm{f} \vert T , D, \Delta)$ of our data given a single non-periodic transit signal of epoch $T$, duration $D$ and depth $\Delta$, for a grid of epochs, durations and depths.
\\\\
To account for correlated noises, we model the light curve $f$ as being drawn from a Gaussian Process such that
$$\bm{f} \sim \mathcal{N}(\bm{w X}, \bm{\Sigma})$$
with mean $\bm{wX}$ (i.e. a linear model of the M explanatory variables with coefficients $\bm{w}$) and covariance $\bm{\Sigma}$. To account for the presence of a single non-periodic transit of epoch $T$ and duration $D$, we compute and append its signal as the last column of the design matrix $\bm{X}$, using the simple transit model from \citealt{protopapas} with a unitary depth. This way, the transit signal is part of the linear model and its depth $\Delta$ can be solved linearly. Under this assumption, the log-likelihood of the data given a single non-periodic transit is analytical
\begin{equation} \label{eq:linear_search_ll}
    \ln p(\bm{f} \vert I) = -\frac{1}{2}(\bm{f}-\bm{wX})^T\bm{\Sigma}^{-1}(\bm{f}-\bm{wX}) -  \frac{1}{2}\vert\bm{\Sigma}\vert - \frac{N}{2}\ln 2
\end{equation}
and can be computed on a grid of epochs and durations, the transit depth being linearly solved for any $(T, D)$. This \textit{linear search} lead to the set of  log-likelihoods\footnote{This notation omits the vector $\bm{w}_{i,j}$ (except its last value $\Delta_{i,j}$) as it is linearly solved for any given value of $(T_i, F_j)$ and irrelevant in what follows.}
$$\ln\mathcal{L} = \{\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})\}_{i, j}$$
were $\Delta_{i,j}$ is the depth linearly solved for a given $(D_i, T_j)$. \autoref{fig:linear_search} shows this likelihood grid computed for the simulated dataset shown in \autoref{fig:principle_dataset}, using the same Gaussian Process and design matrix $\bm{X}$ used to simulate the data (see annexe).

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.75\linewidth]{../figures/principle_linear_search.pdf}
        \caption{Principle and output of the \textit{linear search}. First, a set of durations and depths $\set{T_i, D_j}_{i,j}$ is generated. For each pair of index $(i,j)$, the log likelihood $\ln p(\bm{f} \vert T_i ,D_j, \Delta_{i,j})$ is computed, using the parameters from \autoref{eq:ls} and the expression of \autoref{eq:linear_search_ll}. This process yields the grid of log-likelihoods $\ln\mathcal{L}$ (bottom plot), as well as the $\set{\Delta_{i,j}, \sigma_{i,j}}_{i, j}$ transit depths and errors.}
        \label{fig:linear_search}
    \end{centering}
\end{figure}

To prepare the next step, the corresponding depths $\Delta_{i,j}$ linearly solved for any $(T_i ,D_j)$ are conserved, as well as their associated errors $\sigma_{i,j}$ obtained through the least-square solution
\begin{equation}\label{eq:ls}
    \begin{gathered}
        \bm{w} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}\bm{X}^T\bm{\Sigma}^{-1}\bm{f} \\
        \bm{\sigma} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}
    \end{gathered}
\end{equation}    
where $\Delta_{i,j} = \bm{w}_M$ and $\sigma_{i,j} = \bm{\sigma}_{MM}$

@@

We then need to combine the likelihoods computed from the \textit{linear search} to obtain
\begin{equation}\label{eq:goal}
    p(\bm{f} \vert P, T_0 , D, \Delta)
\end{equation}
For a given transit duration $D$, any combination of $(P, T_0)$ leads to K truly independent transits, for which it is tempting to write
\begin{equation}\label{eq:attempt}
    p(\bm{f} \vert P, T_0 ,D, \Delta) = \prod_k^K p(\bm{f} \vert T_k, D, \Delta_k)
\end{equation}
where $\set{T_k}_k$ are the epochs matching $(T_0, P)$ and $\set{\Delta_k}_k$ the corresponding depths. So that
$$
    \ln p(\bm{f} \vert P, T_0 ,D, \Delta) = \sum_k^K \ln \mathcal{L}_k
$$
This is the joint log-likelihood of transits belonging to the same periodic signal but with varying depths  $\set{\Delta_k}_k$. However, individual transits from a periodic signal cannot be considered independent, and should instead share a common transit depth $\Delta$. We show in \autoref{combining_transits} that there is an analytical expression for the joint log-likelihood of K individual transits with depths and errors $\set{\Delta_k, \sigma_k}_k$ assuming a common depth $\Delta$:
\begin{equation}\label{eq:comb}
    \begin{gathered}
        \ln p(\bm{f} \vert P, T_0 ,D, \Delta) =  \sum_{k}^K \ln \mathcal{L}_k  - \frac{K}{2}\ln(2\pi) - \frac{1}{2}\sum_k^K \left( \frac{\left(\Delta_{k} -
        \bar\Delta\right)^{2}}{\sigma_k^{2} + \sigma^{2}} + \ln{\left(\sigma_k^{2} + \sigma^{2} \right)} \right) \\
        \text{with} \quad  \frac{1}{\sigma^2} = \sum_k^K \frac{1}{\sigma_k^2} \quad \text{and} \quad
        \Delta = \sigma^2 \sum_k^K {\frac{\Delta_k}{\sigma_k^2}}
    \end{gathered}
\end{equation}
While this gives a solution for \autoref{eq:goal}\footnote{Different from the one found in \cite{foreman2016}}, the individual epochs matching $T_0$ and $P$ are not necessarily available in the grid of epochs $\set{T_k}_k$. In \cite{foreman2016} this is solved by using the nearest neighbors in the epochs grid. Instead, to allow the efficient matrix computation of \autoref{eq:comb}, we interpolate the likelihood grid from $\set{T_i}_i$ to a common grid of transit phases $\set{\phi_i}_i$, leading to the \textit{periodic search} log-likelihood
$$\ln\mathcal{P}(P) = \set{\ln p(\bm{f} \vert P, \phi, D)}_{i,j}$$
shown for few periods in \autoref{fig:periodic_search}.b, a notation where $\Delta_{i,j}$ is omitted since being interpolated from the \textit{linear search} using $\phi_i$ and $D_j$, and $T_0 = 0$.

\begin{figure}[H]
    \begin{centering}
        \gridline{\fig{../figures/principle_periodic_0.pdf}{0.72\linewidth}{\vspace{-0.5cm}(a) $\ln \mathcal{L}$}}
        \vspace{-0.6cm}
        \gridline{\fig{../figures/principle_periodic_1.pdf}{0.6\linewidth}{\vspace{-0.5cm}(b) $\mathcal{P}(P)$}}
        \caption{On the example dataset, this figure shows how the \textit{periodic search} works at different periods $P$ (including the true period $P = 0.7$ days). Given $P$ and $T_0=0$, the log-likelihood $\ln\mathcal{L}$ shown in (a) is phase-folded and interpolated onto a common grid of phases shown in (b). As an example, the white lines in (a) mark the edges of each fold for a period of $P=0.5$ days, and the white crosses the epochs $\set{T_k}_{k\in\mathbb{K}}$ matching a particular phase in the grid (reported in (a) and (b)) on which the corresponding $\set{\ln \mathcal{L}_k}_{k\in\mathbb{K}}$ are interpolated and combined using \autoref{eq:comb}. To allow the use of efficient matrix computations, this is done for every duration $\set{D_i}_i$ so that $\mathcal{P}$ is computed on the full grid $\set{D_i, \phi_i}_{i, j}$ at once. We understand from plots in (a) that a different choice of epoch $T_0$ may only shift the results in phase but do not affect the values of $\mathcal{P}$. For this reason, computing $\mathcal{P}$ for $T_0=0$ is sufficient. We also notice how the maximum value of $\mathcal{P}(P=0.35)$ is lower than for $P=0.7$ days, a result of combining the log-likelihoods using \autoref{eq:comb} instead of \autoref{eq:attempt}, in favor of individual transits matching a common depth $\Delta$.}
        \label{fig:periodic_search}
    \end{centering}
\end{figure}

@@

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.65\linewidth]{../figures/principle_Q.pdf}
        \caption{}
        \label{fig:periodogram}
    \end{centering}
\end{figure}

@@

\newpage
\appendix
\section{Light curve simulations}\label{app_principle_simulations}

This paper relies on simulated light curves featuring transit signals, instrumental signals and correlated noises such as stellar variability. Let such a flux $f$ of a star simulated at time $t$ be sampled and aranged in the $(1\times N)$ column-vector $\bm{f}$ associated to the column-vector of times $\bm{t}$. An example of such signal is simulated and shown in \autoref{fig:principle_dataset}.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.8\linewidth]{../figures/principle_dataset_decomposed.pdf}
        \caption{}
        \label{fig:principle_dataset}
    \end{centering}
\end{figure}
In what follows, we describe the models used to simulate the components of this signal, i.e. the periodic transit $\mathcal{T}$, the instrumental systematics $\mathcal{S}$ and the correlated noise $\mathcal{V}$, such that
$$
    f(t) = \textcolor{gray}{\mathcal{T}}(t) %+ \textcolor{Gray}{\mathcal{S}}(t) + \textcolor{Gray}{\mathcal{V}}(t)
$$
computed for $N=2880$ times $t$ corresponding to an observation of 4 days with an exposure time of 2 minutes. 
\subsection{Transit signal}
We simulate the periodic transit signal $\mathcal{T}$ using the simple model described in \cite{protopapas} where a transit of period $P$, epoch $T_0$, duration $D$ and unitary depth observed at time $t$ is given by
\begin{equation}
    \begin{gathered}
        \mathcal{T}_c(t, P, T_0, D) = \frac{1}{2}\tanh\left(c\left(\theta - \frac{1}{2}\right)\right) - \frac{1}{2}\tanh\left(c\left(\theta + \frac{1}{2}\right)\right) \\
        \text{with}\quad\theta = \frac{P}{\pi  D}\sin\left(\frac{\pi(t-T_0)}{P}\right)
    \end{gathered}
\end{equation}
where the dimensionless parameter c controls the roundness of the transit depth ($c\gg1$ corresponding to a box-shaped transit as shown in \autoref{fig:protopapas}). This analytical model is fully empirical but easily differentiable. From this expression, we get the signal of the single non-periodic transit of unitary depth 
\begin{equation}\label{eq:protopapas_single}
    \tau_c(t, T_0, D) = \lim_{P\to \infty} \mathcal{T}(t, P, T_0, D) = -\frac{1}{2}\tan\left(\frac{c}{2}(1-T_0+t)\right) -\frac{1}{2}\tan\left(\frac{c}{2}(1+T_0-t)\right)
\end{equation}
with epoch $T_0$ and duration $D$.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.9\linewidth]{../figures/protopapas.pdf}
        \caption{Simulations of the single non-periodic transit signal $\tau_c(t, T_0, D)$ (\autoref{eq:protopapas_single}) for different values of $c$}
        \label{fig:protopapas}
    \end{centering}
\end{figure}
In our study, we simulate all transit signals with $c=12$. The periodic transit signal $\mathcal{T}$ seen in \autoref{fig:simu} corresponds to $\mathcal{T} = \mathcal{T}_{c=12}(t, P=0.7, T_0=0.2, D=0.05)$, all parameters in unit of \textit{days}.

\subsection{Instrumental signals}
We simulate instrumental signals as a linear model of $M$ explanatory variables arranged in the $(M\times N)$ design matrix $\bm{X}$. Hence,
$$
    \mathcal{S} \sim \bm{\mathcal{S}} = \bm{w}\bm{X}
$$
where the vector $\bm{w}$ are linear coefficients of the model. The simulated flux shown in \autoref{fig:simu} contains a linear model where the $M=4$ columns of the design matrix $\bm{X}$ are given by $\bm{X}_{i} = \bm{t}^i$ (i.e. $\bm{X}$ is the Vandermonde matrix order $3$ of time $t$) and $\bm{w} = [1.0\quad0.0005\quad\min0.0002\quad-0.0005].$

\subsection{Correlated noise}
We generate stellar variability thanks to the \texttt{tinygp}\footnote{\href{https://github.com/dfm/tinygp}{https://github.com/dfm/tinygp}} Python package, providing an implententation of the scalable GP method from \citealt{celerite2} powered by 
\texttt{JAX}\footnote{\href{https://github.com/google/jax}{https://github.com/google/jax}}.
\section{Combining transits}\label{combining_transits}
